### 2016-05-19 Paper_v1.tex
May18

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, California 94305}}
{\par \centering \textbf{Running Title:}~Dynamic model of cell free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Fill me in.

@@

The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).

\begin{enumerate}
	\item{\textbf{First~section}:Description of the model biology}
	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
\end{enumerate}

@@

%As the dimensionality of  increases, the search region gets widened and thus the problem becomes more challenging.
%

@@

The discussion has three (sometimes four) paragraphs:
\begin{enumerate}
	\item{\textbf{First~paragraph}: Present a modified version of the last paragraph of the introduction. In this study, [...]. Taken together, [killer statement]}
	\item{\textbf{Second~paragraph}: Contrast the key findings of the study with other computational/experimental studies}
	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings. }
\end{enumerate}

@@

\subsection*{Formulation and Solution of the Model Equations}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j=1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i=1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij}=0$ indicates metabolite $i$ is not connected with reaction $j$.
Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant.
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right)=\mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predicitons:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.

\subsection*{Global and local sensitivity analysis}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which of the experimentally controllable parameters affected the performance of the reduced order model \citep{SOBOL_METHOD}.
This included the initial conditions of glucose, oxygen, amino acids, and enzymes.
We computed the total sensitivity index of each parameter relative to a performance objective of area under the CAT curve (CAT production).
We established the sampling bounds for each parameter from the value of that parameter in the set used to generate the ensemble.
We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 300 and $d$ = 185, so the total sensitivity indices were computed from
111,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
We conducted a local sensitivity analysis to estimate which of the other model parameters affected performance.
This included the same parameters that were varied in the ensemble: rate constants, saturation constants, and control parameters.
The local sensitivity for each parameter was calculated across a sub-ensemble of 180 parameter sets, randomly chosen from the ensemble of 18,000 sets:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=\frac{p_{ij}}{AUC(p_{ij})}\frac{AUC(p_{ij}+\Delta p_{ij})-AUC(p_{ij})}{\Delta p_{ij}}\qquad{i=1,2,\hdots,\mathcal{E}}\qquad{j=1,2,\hdots,\mathcal{P}} \\
    \Delta p_{ij}=0.001\;p_{ij}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad
\end{split}
\end{equation}
where $\mathcal{E}$ denotes the number of parameter sets in the sub-ensemble, equal to 180, $\mathcal{P}$ denotes the number of parameters, equal to 652, $S_{ij}$ denotes the sensitivity of the $j$th parameter for the $i$th parameter set, $p_{ij}$ denotes the value of the $j$th parameter for the $i$th parameter set, $\Delta p_{ij}$ denotes the perturbation of the $j$th parameter for the $i$th parameter set, equal to 0.1\% of the parameter value, and $AUC()$ denotes the area under the CAT curve.
We then calculated the mean and

\subsection*{Calculation of CAT yield}
The theoritical carbon yield of CAT was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
\begin{equation}\label{eqn:FBA-yield}
\begin{split}
	\max_{\boldsymbol{v}}{} \! \left( v_{obj}=\mathbf{\boldsymbol{c}}^T \boldsymbol{v} \right) \qquad \alpha_i \leq v_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R} \\
	\mathrm{Subject \; to:}	\; \; \mathbf{S}\mathbf{v}=\mathbf{0}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad
\end{split}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{v}$ denotes the unknown flux vector, $\boldsymbol{c}$ denotes the objective selection vector, and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $v_{i}$, respectively.
The stoichiometric matrix was expanded to include the transcription and translation reactions for producing CAT. 
The objective $v_{obj}$ was to maximize the specific rate of CAT formation.
The specific glucose, amino acids and oxygen uptake rates were constrained to allow a maximum flux of 10 mM/hr.
The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
The solution flux vector was used to calculate the theoritical carbon yield of CAT based on carbon consumed (glucose and amino acids).
The carbon yield was formulated as:
\begin{equation}\nonumber
	Yield=\frac{C_{CAT}v_{CAT}}{\sum_{i=1}^{\mathcal{R}}C_{i}v_{i}}
\end{equation}
where $C_{CAT}$ and $C_{i}$ denote the carbon number of CAT and substrate $i$, respectively, $v_{CAT}$ and $v_{i}$ denote the flux of CAT and substrate $i$, respectively,
and $\mathcal{R}$ denotes the number of substrates consumed.

@@

\begin{figure}[ht]
\centering
%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
\caption{A: Central carbon metabolism in the presence of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets. B: Central carbon metabolism in the absence of allosteric control.}
\label{fig:BothCarbon}
\end{figure}

@@

\begin{figure}[ht]
\centering
%\includegraphics[width=1.00\textwidth]{./Figures/SensitivitiesHalfTall.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Metab_sensitivity.pdf}
\caption{Total-order global sensitivities for experimentally controllable initial conditions, including glucose, oxygen, amino acids, and enzymes.}
\label{fig:SensitivitiesHalfTall}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
\label{fig:SensAll}
\end{figure}

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, California 94305}}
{\par \centering \textbf{Running Title:}~Dynamic model of cell free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Fill me in.

@@

%The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
%Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).

%\begin{enumerate}
%	\item{\textbf{First~section}:Description of the model biology}
%	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
%	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
%\end{enumerate}

\subsection*{Estimation of an ensmeble of models}
We constructed an initial model of cell-free \textit{E. coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{?} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset from Swartz and coworkers \cite{?}. 
This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function \cite{?}, and the control term was calculated as the mean of all transfer functions.
%The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of 18,000 parameter sets by minimizing the error between the Swartz dataset and the metabolite concentrations predicted by the model. Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
The ensemble also captures the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

\subsection*{Sensitivity analysis}
We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
This is explained by GTP and the amino acids being reactants for CAT synthesis.
The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
This is because an increase in saturation constant causes a decrease in the corresponding reaction rate \cite{?}.
The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is relatively unimportant to CAT production and metabolism as a whole compared to kinetic effects.

We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
As the sensitivities were total-order, they were guaranteed to be non-negative.
The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
This is all explained by GTP and amino acids being reactants for CAT synthesis.
While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.

\subsection*{Calculation of CAT yield}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
Arginine and glutamate were excluded due to not being present in the Swartz dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
Thus, we showed that our experimental dataset and best-fit parameter set were producing CAT at 44.5\% and 44.9\% of the theoretical maximum, respectively.
This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.

@@

@@

The discussion has three (sometimes four) paragraphs:
\begin{enumerate}
	\item{\textbf{First~paragraph}: Present a modified version of the last paragraph of the introduction. In this study, [...]. Taken together, [killer statement]}
	\item{\textbf{Second~paragraph}: Contrast the key findings of the study with other computational/experimental studies}
	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings.}
\end{enumerate}

@@

\subsection*{Formulation and solution of the model equations}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j=1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i=1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij}=0$ indicates metabolite $i$ is not connected with reaction $j$.
Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant.
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right)=\mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.

\subsection*{Global and local sensitivity analysis}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which of the experimentally controllable parameters affected the performance of the reduced order model \citep{SOBOL_METHOD}.
This included the initial conditions of glucose, oxygen, amino acids, and enzymes.
We computed the total sensitivity index of each parameter relative to a performance objective of area under the CAT curve (CAT production).
We established the sampling bounds for each parameter from the value of that parameter in the set used to generate the ensemble.
We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 300 and $d$ = 185, so the total sensitivity indices were computed from
111,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
We conducted a local sensitivity analysis to estimate which of the other model parameters affected performance.
This included the same parameters that were varied in the ensemble: rate constants, saturation constants, and control parameters.
The local sensitivity for each parameter was calculated across a sub-ensemble of 180 parameter sets, randomly chosen from the ensemble of 18,000 sets:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=\frac{p_{ij}}{AUC(p_{ij})}\frac{AUC(p_{ij}+\Delta p_{ij})-AUC(p_{ij})}{\Delta p_{ij}}\qquad{i=1,2,\hdots,\mathcal{E}}\qquad{j=1,2,\hdots,\mathcal{P}} \\
    \Delta p_{ij}=0.001\;p_{ij}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad
\end{split}
\end{equation}
where $\mathcal{E}$ denotes the number of parameter sets in the sub-ensemble, equal to 180, $\mathcal{P}$ denotes the number of parameters, equal to 652, $S_{ij}$ denotes the sensitivity of the $j$th parameter for the $i$th parameter set, $p_{ij}$ denotes the value of the $j$th parameter for the $i$th parameter set, $\Delta p_{ij}$ denotes the perturbation of the $j$th parameter for the $i$th parameter set, equal to 0.1\% of the parameter value, and $AUC()$ denotes the area under the CAT curve.
We then calculated the mean and standard error of each local sensitivity across the sub-ensemble of 180 sets.

\subsection*{Calculation of CAT yield}
The yield on CAT production was calculated for three cases: the experimental data, the best-fit parameter set, and a theoretical maximum yield.
In each case the yield was formulated as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
\begin{equation}\label{eqn:yield-definition}
	Yield=\frac{\Delta CAT\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\;C_{m_i}}
\end{equation}
where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i$th reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i$th reactant.
Because no data was available for arginine or glutamate, these reactants were left out of all three calculations.
In the experimental case and the best-fit set case, yield was calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
\begin{equation}\label{eqn:FBA}
\begin{split}
	\max_{\boldsymbol{v}}{} \! \left( v_{obj}=\mathbf{\boldsymbol{c}}^T \boldsymbol{v} \right) \qquad \alpha_i \leq v_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R} \\
	\mathrm{Subject \; to:}	\; \; \mathbf{S}\mathbf{v}=\mathbf{0}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad
\end{split}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{v}$ denotes the unknown flux vector, $\boldsymbol{c}$ denotes the objective selection vector, and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $v_{i}$, respectively.
The stoichiometric matrix was expanded to include the transcription and translation reactions for producing CAT. 
The objective $v_{obj}$ was to maximize the specific rate of CAT formation.
The specific glucose, amino acids and oxygen uptake rates were constrained to allow a maximum flux of 10 mM/hr.
The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
The solution flux vector was used to calculate the theoretical carbon yield of CAT, reformulated in terms of flux:
\begin{equation}\label{eqn:yield}
	Yield=\frac{v_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(v_{i},0)\;C_{i}}
\end{equation}
where $v_{CAT}$ denotes the CAT export flux and $v_{i}$ denotes the import flux of the $i$th substrate.

@@

\begin{figure}[ht]
\centering
%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
\label{fig:BothCarbon}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
\label{fig:LocalSensitivity}
\end{figure}

\begin{figure}[ht]
\centering
%\includegraphics[width=1.00\textwidth]{./Figures/SensitivitiesHalfTall.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
\includegraphics[width=1.00\textwidth]{./Figures/Metab_sensitivity.pdf}
\caption{Total-order global sensitivities for experimentally controllable initial conditions, including glucose, oxygen, amino acids, and enzymes.}
\label{fig:GlobalSensitivity}
\end{figure}
